Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-21 aistats.tex
Add self-concordance erratum

@@

\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
\part{Appendix} % Start the appendix part
\parttoc % Insert the appendix TOC

@@

In this section, we prove that for a Gaussian, the entropy is self-concordant while the log-partition function is not.
We also give complementary illustrations on the Gaussian mirror map and MLE.

The Gaussian log-partition function and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; ,
}
where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
These functions and their domains are represented in Figure \TODO{}.
We provide definitions of self-concordance in \cref{app:self-concordant}.

@@

\paragraph{Log-partition is Self-Concordant}
Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(\alpha \ \ -1)^\top$. 
For varying $\nat_1, \alpha$, this covers the whole range of possible straight lines in the 2D plane, except lines of constant $\nat_2$ (horizontal).
But on lines with $\nat_2=\cst$,   the log-partition function $A$ is simply a quadratic, so it is self-concordant.
Then, up to constants,
\alignn{
g(t) := 2A(w+tv) 
= \frac{(\nat_1 + t \alpha)^2}{2 t} - \log(t) + \cst
= \frac{a}{t}  + b t - \log(t) + \cst
}
where $a=\frac{\nat_1^2}{2} \geq 0$, and $b = \frac{\alpha^2}{2}$.
$g$ is well defined for $t>0$.
The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
Linear combinations with positive coefficients  preserves self-concordance \citep[theorem 4.1.1]{nesterov2003introductory} -- e.g. self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant. 
So let us prove it.
The derivatives of $g$ are
\alignn{
g'(t) &= -\frac{a}{t^2} - \frac{1}{t} + b \; ,\\
g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\
g'''(t) &= -\frac{6a}{t^4} - \frac{2}{t^3} = \frac{-2}{t^3}\paren{\frac{3a}{t} + 1} \;.
}
Then the definition of self-concordance reads
\alignn{
\abs{g'''(t)} &\leq 2 g''(t)^{\frac{3}{2}} \\
\iff \frac{3a}{t} + 1 &\leq \paren{\frac{2a}{t} + 1}^{\frac{3}{2}} \; .
}
Writing $y = \frac{a}{t} \geq 0$, and taking the square of this expression yields
\alignn{
 (3y + 1)^2 &\leq (2y + 1)^3 \\
 \iff 9 y ^2 + 6y + 1 &\leq 8 y^3  + 12 y^2 + 6y +1 \\
\iff 0 &\leq y^2(8y + 3)
}
which is true since $y\geq 0$.
The log-partition is self concordant.

@@

\TODO{Damien : review self-concordance + prove proposition 1 + graph to explain $x<0.21$.}

@@

@@

\subsection{Self-concordance and Convex Conjugates}
\begin{proposition}
	\label{prop:conjugateConcordance}
	Let $f:\real \rightarrow \real$ be a self-concordant function of Legendre type. Then its convex conjugate $f^*$ is also self-concordant.
\end{proposition}
\begin{proof}
When $f$ is of Legendre type then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ has inverse gadients
\begin{align}
    \nabla f \circ \nabla f^*  &= \Id \\
     \nabla f^* \circ \nabla f &= \Id
\end{align}
where $\Id$ is the identity function on the relevant domain. 
Deriving this equality proves that Hessians are inverse of each other 
\begin{align}
    \nabla^2 f(x) \nabla^2 f^*(x^*) = \mI
    \label{eq:conjugateHessian}
\end{align}
when evaluated at conjugate pairs of points $(x,x^*)$ -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
These Hessians are never $0$.
In 1 dimension, this inequality is more simply written 
\aligns{f''(x) f^{* \prime \prime}(x^*) = 1}
Until now, this should be familiar to the reader.
The interesting surprise is that self-concordance appears when we derive again this equality. 
\begin{align}
    &1= f''(x) f^{* \prime \prime}(f'(x)), \forall x \\
    \implies
    &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
    \iff
    &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\
    \implies
    &0 = \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  f^{* \prime \prime}(x^*)^{-\half}
    + f''(x)^{\half} \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}} \\
    \implies
    &0= \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}}
\end{align}
where to reach the last line, we used~\eqref{eq:conjugateHessian} again.
We see that for a pair of conjugate functions, the self-concordance ratios are opposite of each other, which concludes the proof.
\end{proof}

>>>>

% To make colored boxes
\usepackage[framemethod=TikZ]{mdframed}
\newmdenv[
	roundcorner=10pt,
	backgroundcolor=Red!30,
]{important}

@@

\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
\part{Appendix} % Start the appendix part
\begin{important}
	{\bf Erratum:}  contrary to what is written in the main text, the log-partition of a Gaussian \emph{is} self-concordant. See \cref{app:gaussian} for details. 
\end{important}
\parttoc % Insert the appendix TOC

@@

\begin{important}
{\bf Erratum:} There was an error in our derivations. In the main text, we claimed that the log-partition of a Gaussian was not self-concordant. We will now prove that the opposite is true.
\end{important}

In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
%We also give complementary illustrations on the Gaussian mirror map and MLE.

The Gaussian log-partition function and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; ,
}
where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
%These functions and their domains are represented in Figure \TODO{}.
We provide definitions of self-concordance in \cref{app:self-concordant}.

@@

\paragraph{Log-partition is Self-Concordant}
As proved in \citet{nesterov1994interior} self-concordance is preserved by Fenchel conjugacy.
Since $\conj$ is self-concordant, $\logpart$ as well.
For a more accessible reference, see also \citet[Prop.~6]{sun2019generalized}.

@@

@@

{\bf Correction:} In the main text, we claimed that the Fenchel conjugate of a 1 dimensional function is also self-conjugate. 
Actually, this is also true in higher dimension, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.

@@

\subsection{Self-concordance and Convex Conjugates}
\begin{proposition}
	\label{prop:conjugateConcordance}
	Let $f:\real \rightarrow \real$ be a self-concordant function of Legendre type. Then its convex conjugate $f^*$ is also self-concordant.
\end{proposition}
\begin{proof}
When $f$ is of Legendre type then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ has inverse gadients
\begin{align}
    \nabla f \circ \nabla f^*  &= \Id \\
     \nabla f^* \circ \nabla f &= \Id
\end{align}
where $\Id$ is the identity function on the relevant domain. 
Deriving this equality proves that Hessians are inverse of each other 
\begin{align}
    \nabla^2 f(x) \nabla^2 f^*(x^*) = \mI
    \label{eq:conjugateHessian}
\end{align}
when evaluated at conjugate pairs of points $(x,x^*)$ -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
These Hessians are never $0$.
In 1 dimension, this inequality is more simply written 
\aligns{f''(x) f^{* \prime \prime}(x^*) = 1}
Until now, this should be familiar to the reader.
The interesting surprise is that self-concordance appears when we derive again this equality. 
\begin{align}
    &1= f''(x) f^{* \prime \prime}(f'(x)), \forall x \\
    \implies
    &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
    \implies
    &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\
    \implies
    &0 = \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  f^{* \prime \prime}(x^*)^{-\half}
    + f''(x)^{\half} \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}} \\
    \implies
    &0= \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}}
\end{align}
where to reach the last line, we used~\eqref{eq:conjugateHessian} again.
We see that for a pair of conjugate functions, the self-concordance ratios are opposite of each other, which concludes the proof.
\end{proof}
